---
title: "Aggregating data"
author: "Esko & Markus"
date: "26.08.2016"
output: html_document
---
  
In the previous example, we compared the performance of children and adults in the learning game and also investigated the association between the game performance and the working memory test. We had a neatly formatted data, where each participant formed a case and a couple of variables summarized the participant's performance. 

However, this is not the original data we collected. The original data is actually in a completely different format. I used a script written in R to transform it into an intermediate format, in which each row corresponds to an individual response. This data is available in file kr_gameraw.txt.

```{r}

game.raw <- read.table("kr_gameraw.txt", header=T)
head(game.raw)
```

The variables in the dataset are as follows:
- player_id : the participant id
- age_grp: A = adult participant, C = child participant
- gender ; Nainen = female, Mies = male
- clip_id : Specifies the video clip that was watched. There were a total of 30 clips in random order.
- video_id : video_id tells the position of the video watched in the sequence. Values range from 0 (the first) to 29 (the last).
- box_id : Identifies the locations presented for the clip. Remember, the task is to select those locations where there was a road user at the time when the video was masked
- correctRejection, falseAlarm, miss, hit : TRUE or FALSE depending on the answer and location. 

Question: How to calculate the proportion of hits, misses etc.
-----------------------------------------------------------------

We had values Phit, Pmiss, PfalseAlarm, PcorrectRejection etc. which we used in the calculations. Basically, each of these represent the proportion of TRUEs in the for each participant.

First, we need to understand how R deals with boolean values (TRUE, FALSE). R treats TRUE essentially equivalent to 1 and FALSE as equivalent to 0. 

```{r}

TRUE == 1
FALSE == 0
TRUE == 0
FALSE == 1
```
This happens also with other operators and functions:

```{r}

TRUE + TRUE
TRUE - TRUE
(TRUE + FALSE + TRUE) / 3

mean(c(TRUE, FALSE, TRUE))
sum(c(TRUE, FALSE, TRUE))
```

This means that it is possible to calculate the proportion of TRUEs in a vector by simply using the function _mean_. 

```{r}
tfs <- c(TRUE, TRUE, FALSE, TRUE, FALSE)
mean(tfs)

```

In other words, we could calculate the proportion of hits among all the answers in the following way:

```{r}
mean(game.raw$hit)

```
But we want to calculate this (and other values) for each participant and store the results into a new variable in a new data frame. This is called aggregation, and there is a very useful package called _plyr_ to do this kind of work. 


Plyr: split, apply and combine
-------------------------------------------------
  
We have already seen how to calculate new values based on existing variables. We have also seen how to create a subsets of data. Next we combine these two ideas to have an extremely useful and powerful data manipulation techinique: split, apply and combine. After you master this, you will survive most of the data manipulation problems you will encounter. 

Splitting means that we first divide the data frame by some variable(s), which are typically of type _factor_. Then we apply one or more function(s) to calculate new variables based on the subsets of data. Then we combine the calculated variables together with the values which were used as a basis for splitting. 

The package plyr combines these three operations into a single command. 

```{r}
require(plyr)

params <- ddply(game.raw, .(player_id), summarise,
                  Phit = mean(hit))
head(params)

```

We got a new data frame which contains the player_id and the proportion of hits. 

The call has the following parts:

- ddply : The function called, Here dd refers to the fact the this function takes a data frame as input and produces a new dataframe as output. 
- game.raw : The first argument specifies the data frame given as input   
- .(player_id) : Specifies by which variable the data frame is to be splitted to pieces. The variable should be of type _factor_. If it is not, it will be converted into one. 
- summarise : This is a function that will be called for each piece. 
- Phit = mean(hit) : This tells the summarise function that we want to calculate the mean of the values in the variable _hit_ and create a new variable called _Phit_ to store the result. 

Note that the function call could be written _summarize_ instead of _summarise_... Sometimes there is a possibility for a name clash with package Hmisc, which also conatins a function called _summarize_

More than one grouping variable
-------------------------------------------

In case we would like to retain the information about the age group for each participant, it is possible by just inserting age_grp as another grouping variable. 

```{r}
params <- ddply(game.raw, .(age_grp, player_id), summarise,
                  Phit = mean(hit))

head(params)
```
Here adding age_grp variable to the grouping factors does not affect the Phit values calculated, because each player_id belongs to either to adult or children group. It does not matter weather age_grp is inserted before or after player_id. 

Now we can actually calculate the dataset we previously used by adding a couple more calculations.
```{r}
params <- ddply(game.raw, .(age_grp, gender, player_id), summarise,
                  Phit = mean(hit),
                  Pmiss = mean(miss),
                  PfalseAlarm = mean(falseAlarm),
                  PcorrectRejection = mean(correctRejection))
head(params)

write.table(params, file="kr_game.txt", row.names=F)

```

Calculating descriptive statistics using plyr
--------------------------------------------------
As we now have a single observation for each participant, we can get some descriptive statistics for the age groups using the same command again. 
```{r}

desc <- ddply(params, .(age_grp), summarise, 
                      M = mean(Phit),
                      SD = sd(Phit),
                      N = length(Phit))
print(desc)

```

We could use the function _kable_ from the package _knitr_ to render this in a nicer way.  
```{r}
require(knitr)
kable(desc, digits=2)
```

And if we would like to change the proportions to percents, we can just multiply the numbers by 100. 
```{r}

desc <- ddply(params, .(age_grp), summarise, 
                      M = mean(100*Phit),
                      SD = sd(100*Phit),
                      N = length(Phit))
kable(desc, digits=0)
```


