---
title: "Reading and writing data"
output: html_document
---


This is partly based an excellent tutorial [R for SAS and SPSS users](https://science.nature.nps.gov/im/datamgmt/statistics/r/documents/r_for_sas_spss_users.pdf)


Let's start with a simple example.We have a dataset in a text file. The variables are separated with tabs (empty space). Variables names are on the first row. 

```{r}

apgar <- read.table("apgar.txt")
head(apgar) # preview the start (i.e. head) of the data frame

```
That was not complicated. However, there are a couple things you need to pay attention to. 

*Working directory*

The filenames are resolved relative to the current working directory of your R console. If the file is in the working directory it can be found with the filename. If not, you either need to copy the file to your working directory or give also the path to the file. 

But what is your working directory? Use function _getwd_ to find it out.  

```{r}
getwd()

```


Sometimes you may need to change your current working directory. The easiest way is to you the Files-browser in Rstudio, go to the right directory, and the click More >> Set As Working Directory. You can set the working directory with the function _setwd_.

```{r}
setwd("~/github/kickstartr/2016aug")  # write the right path 

```

When you create a new project in RStuduio it sets the working directory to be the same as the directory where the project file is saved. When you load the project, the working directory is automatically restored. 


*Separator used* 

By default, _read.table_ uses any whitespace character to separate different variables. However, the separator can be something else in the data file, like comma. In apgar.csv the same data is in CSV format (CSV = comma separated values), and the simple call to read.table fails. 

The problem is easily corrected by specifying the right separator with _sep=','_. However, in order to get also the variable names correctly interpreted, we need to give the parameter _header=TRUE_ also.
```{r}
apgar.notyet <- read.table("apgar.csv", sep=",")
head(apgar.notyet)

apgar.ok <- read.table("apgar.csv", sep=",", header=TRUE)
head(apgar.ok)
```

Alternatively, you can use function read.delim, read.csv etc. They are just shortcuts to read.table with slightly different default values for the parameters like separator. 

*Decimals are marked with comma, not with point*

In Finnish, the convention is to use decimal comma, not point. By default SPSS, Excel etc. detect the locale settings of the operating systems and adapt their notation accordingy. This creates sometimes problems with R and many other programming environments, which typically follow the English convention of decimal point. 


```{r}
# Separator is here ;W
apgar.dc <- read.table("apgar_decimalcomma.txt", sep=";")
head(apgar.dc)
```

At first, everything seems to be fine, before you try to use some of the decimal values...

```{r}
apgar.dc$lgincome * 2
```

... which has been interpreted as string, and then converted to factors by _read.table_.

```{r}
class(apgar.dc$lgincome)
```

This can solved by specifying the decimal point character. 

```{r}
apgar.dc2 <- read.table("apgar_decimalcomma.txt", dec=",", sep=";")
head(apgar.dc2)
```


Reading SPSS files
------------------------
R can read for example SPSS-files directly, using package foreign

```{r warning=FALSE}
library(foreign)
apgar <- read.spss("apgar.sav",to.data.frame=TRUE)

View(apgar)
```


Writing data
=================================
All the above presented text data files were actually created from the aforementioned apgar.sav file. The following functions calls were used. 

```{r}
write.table(apgar, "apgar.txt", sep="\t")
write.csv(apgar, "apgar.csv", row.names=F) # by default separated with comma
write.table(apgar, "apgar_decimalcomma.txt", sep=";", dec=',')

```

