---
title: "Representing data and making simple calculations"
date: "24.08.2016"
output: html_document
---

Computers are stupid. If we want to get them to make data analysis work for us, we need to tell them exactly 1) what is the data they are supposed to use, 2) what are the tools to calculate new values from the existing values, and 3) at which order these calculations are performed on which data. 

_Objects_ are the way to represent data in a programming language like R so that a computer can use it. 

_Operators_ are tools to calculate new values from existing values. 

_Scripts_ define algorithms, that is, instructions for using the right objects and operators at the right time. 


Objects
==========================

For getting started with R, you need to know grasp the idea of two kinds of objects: _vectors_ and _data frames_. 

Vector
-------------

_Vector_ is an ordered collection of data, like numbers or strings. 

When you enter text to the R console, the interpreted will parse your input, and create a vector. The vectors "know" what kind of data they have in them.

Try to run to enter the input below to R console.
```{r results="hide"}
# creates a vector, which contains only one number, and then prints the contents of that vector
3    

# creates a vector with three numbers, and then prints the contents of that vector
c(7, 6, 3)
```


Vectors have length and type
```{r results="hide"}
# the vectors have type
class(3)
class(c(7, 6, 3))

# and length
length(3)
length(c(7, 6, 3))
```

Vector may contain also strings, not only numbers
```{r results="hide"}
# creates object of type character
"Vector"
c("Vector", "says", "hello!")

# beside the type, everything is same as with numbers
class("Vector")
class(c("Vector", "says", "hello!"))

length("Vector")
length(c("Vector", "says", "hello!"))
```

[More about vectors in R language defition](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Vector-objects)

Variables 
-----------------------

In the above example, we created vectors on the fly. However, typically we want to create some data (or read it from a file) and then store it to a variable for later use. 

```{r}
foo <- c(2, 4, 6, 8)
bar <- "Hei hoi!"
```

With variables we have a new way to use vectors which has more than one member. We can access each member by using a its index. It is also possible change individual members. 

```{r}
foo[2]
foo[2] <- 500
foo
```

You can use variables also in creation of new vectors.
```{r}
thesisWorker <- 0
researchAssistant <- 1700
doctoralStudent <- 2200
postdoc <- 3200
professor <- 5500
rector <- 25000

salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)

```


Operators
========================

Objects represent data and variables can represent them in algorithms.

With operators new objects are created from existing objects. 

However, at first there is nothing that exciting if you remember anything about elementary school mathematics. 

```{r results="hide"}
# Let's create objects and assign them to variables.
a <- 4
b <- 6
c <- -3

a + b
a - b
a * b
a / b

# NB. Order or calculation follows the ordinary mathematical rules
a + b * c    
(a + b) * c  
a + (b * c) 

```

These we are vectors which had a length of 1. Everything happens pretty much the same way for vectors which are longer. Every operator is applied for every element, separately. It is said the operator is applied element-wise. 

```{r}
many_numbers <- c(1, 2, 3, 4, 5)
many_numbers * 2

```

Comparative operators 
--------------------------------

In the above, we used comparator operator >. They can be used to compare things, and the results will be either TRUE or FALSE. 


```{r results="hide"}

5 > 2   # is greater than
5 < 2   # is smaller than
5 == 2  # is equal to
5 != 2  # is not equal to
5 >= 5   # is greater or equal
5 <= 5   # is smaller or equal 

!(5 == 2)  # ! alone is negation

```

[More about operators in R language definition](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Operators)




Scripts or algorithms
========================

The power of any programming language is on the possibility to describe the algorithm (= what will be computed) regardless of the actual data. For this we need variables, which can represent the actual objects in the algorithm consisting of steps where operators are used to calculate new values from existing values.  

```{r}

salaryBeforeTaxes <- 2734.51

employerCosts <- salaryBeforeTaxes * 1.56 

print(employerCosts)

```

Here the R console first creates a vector of type numbers with one member (2734.51). Then it is assigned to the variable salaryBeforeTaxes with operator <-. You can thing a variable a box where you can but data. 


More objects: Data frames
===================================

Data frame can be tought as a collection of vectors, so that the vectors form a two-dimensional table.  

Two-dimensional table it the most common way to represent data, at least in behavioural sciences. On the rows, there are observations and on the columns, there are different variables which describe different aspects of these observations.

Data frames can be constructed by first creating some vectors (of the same length), and then combining them with function _data.frame_.

```{r}

# create vectors
position <- c("Thesis worker", 
               "Research assistant",
               "Doctoral student",
               "Postdoc", 
               "Professor", 
               "Rector")
salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)
workingYears <- c(0, 1, 3, 8, 20, 30)
citations <- c(0, 
               1,
               8,
               30,
               7784,
               2021)

# create a data frame
academics <- data.frame(position, salaryBeforeTaxes, citations, workingYears)

```

We can view this nice data frame either printing it to the console or using View. 

```{r}
academics # print all
View(academics) # view in a separate tab/window
head(academics) # print only the first rows

```




Accessing the data in data frames
-----------------------------------------

Vectors (variables) within data frames can be easily accessed and used with $.  

```{r}

academics$salaryBeforeTaxes
```

Using function names, it is also possible to get all the variables names as a vector. 

```{r}
# get the names as a vector
names(academics)  

```
This can be also used to change them. 
```{r}
# set the names from a vector
academics.cryptic <- academics # make a copy of the data frame first

names(academics.cryptic)

# Use short cryptic variable names and you will be viewed as
# - super-intelligent by all non-programmers 
# - super-stupid by all who know how to program
names(academics.cryptic) <- c("pos", "slr", "cit", "wyr")   
# Remember that your scripts are the best (and typically only) documentation
# for your data-analysis. Make them readable and accessible! 
names(academics.cryptic)
```

Because they were first creates as vectors, they behave like vectors also when they are part of a data frame. 
```{r}
academics$salaryBeforeTaxes * 1.56
```

Extra stuff: Attach and detach
-----------------------
Notation can be simplified by giving a certain data frame a privileged status using attach:

```{r}
academics$citationsPerYear <- NULL # remove it so that it is newly created

attach(academics)

citationsPerYear <- citations / workingYears

detach(academics)

```



Creating new variables to data frames 
----------------------------------------------------

Often we need to calculate something from the existing variables in the data frame and then include them to the data frame. 

```{r}
academics$costs <- academics$salaryBeforeTaxes * 1.56

academics$citationsPerYear <- academics$citations / academics$workingYears

academics

```

```{r}
academics$citatonsPerCitations <- academics$citations / academics$citations
academics

# If we made a mistake in creating the new variable, we can "cancel" creating it. 
academics$citationsPerCitations <- NULL

# The command assigned "emptiness" as the value of the variable, and see, it is gone!
academics
```

Extra stuff: Other way to create new variables
-----------------------------------------------

There are also other ways for that

```{r}
academics$citationsPerYear <- NULL

academics <- transform(academics,
          citationsPerYear = citations / workingYears)
academics
```










