---
title: "rmANOVA"
author: "Esko"
date: "01.07.2016"
output: html_document
---


Dataset
----------------

We start with the raw game data from the previous sections.

```{r}
game.raw <- read.table("kr_gameraw.txt", header=T)
head(game.raw)
```

The data was about learning game, where each player player 30 video clips. 

Question: Did players' performance improve while playing the game? 
-------------------------------------------------------------------

A very straigforward way to examine this question would be to divide the 30 videos clips into three sets, each containing 10 clips. Then we could calculate -- that is, aggregate -- the _Phit_ value by each set and player. 

Recoding video\_id into a new variable, _set_
--------------------------
There is the variable _video\_id_, which codes the order of the videos from 0 to 29. We can use it to form a new variable called _set_. 

```{r}
# Using a quite easy to understand recode function from package car
require(car)
game.raw$set <- recode(game.raw$video_id, "0:9=1; 10:19=2; 20:29=3")

# The same result can be obtained by doing some math and using the function "floor" to round downwards. 
# E.g. 7 / 10 = 0.7; floor(0.7) = 0; 0 + 1 = 1

game.raw$set2 <- floor(game.raw$video_id / 10) + 1

# The results are identical. It is up to you which one to use.  
identical(game.raw$set, game.raw$set2)
```

Finally, it is a good idea to convert the variable _set_ explicitly into an ordered factor.

```{r}
game.raw$set <- ordered(game.raw$set)
```


Aggregation by player and set
--------------------------------
```{r}
require(plyr)
params <- ddply(game.raw, .(age_grp, player_id, set), summarize,
                  Phit = mean(hit))

head(params)
```

We could also include more parameters in the calculation.

```{r}
params <- ddply(game.raw, .(age_grp, player_id, set), summarize,
                  Phit = mean(hit),
                  PcorrectRejection = mean(correctRejection),
                  accuracy = Phit + PcorrectRejection)
head(params)
```

We calculate both _Phit_ and _PcorrectRejection_. Then we calculate _accuracy_ by adding them. This is possible already here, inside the ddply-call, because the variable _Phit_ and _PcorrectRejection_ are created before accuracy. 

To answer the analysis question, it makes sense to first produce a plot and then do a repeated measures ANOVA. In the next section we will use ggplot2 to make some fancier plots, but let's just stick to the R's built-in plotting tools for now.  

```{r}
# boxplot
boxplot(accuracy ~ set + age_grp, params)

# and this could be used to summarise the effects using means
interaction.plot(params$set, params$age_grp, params$accuracy)
```

Repeated-measures ANOVA
-------------------------

Calculating a repeated measures ANOVAs in R is a bit complicated using the built-in functions because you need to understand what you are doing. There I recommed using package ez. 


```{r}
##################################################################################
# Before calculating anovas using R, set the default contrasts to orthogonal, 
# if you want to have results comparable to SPSS. 
#
# Especially, if you want to use type III Sum of Squares like in SPSS by default, 
# you MUST use orthogonal contrasts! Otherwise your results may end up wrong. 
##################################################################################
options(contrasts = c("contr.sum", "contr.poly")) 

require(ez)
fm <- ezANOVA(params, 
        dv=accuracy,
        wid=player_id,
        within = .(set),
        between = .(age_grp),
        type=3) # we ask for type III Sum of Squares
# If you didn't convert the variable "set" into a factor, you will run into problems here, because ezANOVA # will then treat it as numeric, which will produce incorrect results

print(fm)
```

How about the post-hoc testing? Ask the ezANOVA to provide the linear model also, and then feed it to lsmeans. 

```{r}
# I repeat this here: Never forget to use orthogonal contrasts with type III Sum of Squares! 

options(contrasts = c("contr.sum", "contr.poly")) 

fm <- ezANOVA(params, 
        dv=accuracy,
        wid=player_id,
        within = .(set),
        between = .(age_grp),
        type=3,
        return_aov=T) # get the model

require(lsmeans)

# pairwise post-hocs
lsmeans(fm$aov, pairwise ~ set)
# with summary, you can get e.g. bonferroni or holm adjusted pairwise comparisons
summary(lsmeans(fm$aov, pairwise ~ set), adjust="bonferroni") 
summary(lsmeans(fm$aov, pairwise ~ set), adjust="holm") 


# polynomial contrast
lsmeans(fm$aov, poly ~ set)
```

Because the main effect of set was significant (and the interaction was not), and the polynomial contrasts indicate positive linear effect, we can conclude that the players are learning.

```{r}
# It is possible to investigate interactions also, but remember that 
# F test for interaction was not significant. So there are for 
# illustration only.

# a bit complicated and makes too many tests
lsmeans(fm$aov, pairwise ~ set*age_grp)

# maybe better
lsmeans(fm$aov, pairwise ~ set | age_grp)
lsmeans(fm$aov, poly ~ set | age_grp)
```