---
title: "Select observations"
author: "Markus"
date: "08 Jul 2016"
output: html_document
---

Selecting observations from a data frame
====================================================

```{r results = "hide", echo=FALSE}

thesisWorker <- 0
researchAssistant <- 1700
doctoralStudent <- 2200
postdoc <- 3200
professor <- 5500
rector <- 25000

salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)

salaryBeforeTaxes 

employerCosts <- salaryBeforeTaxes * 1.56 
employerCosts

position <- c("Thesis worker", 
              "Research assistant",
              "Doctoral student",
              "Postdoc", 
              "Professor", 
              "Rector")

workingYears <- c(0, 1, 3, 8, 20, 30)

citations <- c(0, 
               1,
               8,
               30,
               7784,
               2021)
academics <- data.frame(position, salaryBeforeTaxes, citations, workingYears)

# bachelor <- c(1,3,4,4,5,5)
# master <- c(1,2,4,5,5,5)
# doctor <- c(1,1,1,3,5,5)
# noeasy <- c(5,5,3,2,1,5)
# 
# thesis.questionnaire <- data.frame(bachelor,master,doctor,noeasy)
# thesis.questionnaire$position <- position
# thesis.questionnaire.explanation <- list()
# thesis.questionnaire.explanation$bachelor <- "Writing a bachelor's thesis is easy"
# thesis.questionnaire.explanation$master <- "Writing a master's thesis is easy"
# thesis.questionnaire.explanation$doctor <- "Writing a doctoral thesis is easy"
# thesis.questionnaire.explanation$noeasy <- "Writing any academic thesis is difficult"
# thesis.questionnaire.explanation$scale <- "1 = Disagree completely, 2 = Disagree somewhat, 3 = Neither disagree or agree, 4 = Agree somewhat, 5 = Agree completely"
# thesis.questionnaire.explanation
# 
# # AHA! Rector's lying! His answers to the questionnaire instrument are not congruent
# 
# academics2 <- cbind(academics,thesis.questionnaire[,1:4])
```

Previously, we created the dataset academics with the questionnaire responses.
Let's modify the data a bit further

```{r}
# Let's give the academics genders:

academics$gender <- c("M","M","F","M","F","M")

# The names of the observations are stored in the row names vector that can be accessed using:

row.names(academics)

# As you can see, the names are enclosed in quotes, so they are of type "character"
# We could also give actual names to the observations, e.g.

row.names(academics)[1] <- "Cosmo Kiveskoski"
row.names(academics)[2] <- "Kalja Naulapää"
row.names(academics)[3] <- "Miina Kenttä"
row.names(academics)[4] <- "Heppu Kaalinpää"
row.names(academics)[5] <- "Merikeiju Jumalankiitos"
row.names(academics)[6] <- "Meirami Pervonkallio"

academics

# Oops! There's a mistake in the gender of the rector! Let's correct:

academics["Meirami Pervonkallio","gender"] <- "F"
academics

#Names can be changed back to numbers by saying:
row.names(academics) <- as.character(1:6)

#The result is equivalent to just saying:
row.names(academics) <- 1:6
row.names(academics)
# But as we can see, the row names are enclosed in quotes, so they are of the type "character"
# So R implicitly converts the numbers to characters for row names, trying to guess what we want
```

Sometimes it's necessary to create subsets of the data, to select only certain observations.

```{r}
# So how to select, e.g. only males?
# We can do this by indexing the data frame as follows:

academics[academics$gender == "M",]

# The latter part actually creates a boolean vector (a vector of TRUE/FALSE values):

academics$gender == "M"
```

It's good to note that R is case-sensitive:
```{r}

academics$gender == "m"

# So we are asking the program to select the observations for which the logical expression is TRUE
```

Even though subsetting data frames using indices is a necessary skill to learn while using R,
there is also a perhaps more intuitive command that can be used. It's called *subset* and it's a good option as long as you are not writing functions of your own. It works as follows:

```{r}
subset(academics, salaryBeforeTaxes > 3000)
```

Here we chose the observations for which the logical test is TRUE. The *subset* command can also be used for selecting only certain variables using the option *select*

```{r}
subset(academics,salaryBeforeTaxes > 3000, select = 3:4)
```

Then, a couple of exercises:

How could we ask for the summary of all variables for FEMALES?

Are the salaries of the females and males in this representative sample of academics equal?

Back to Indexing
---------------------

Couple of more details of related to using indices:

```{r echo=FALSE, results="hide"}
academics[ c(3,4,5,6), ] # is equal to:
academics[ c(3:6), ] # is equal to:
```
```{r}
academics[ -c(1:2), ]
```

To understand the negative indices, you can think of Venn diagrams and the negation of a set.

We can also use our brand new brilliant row names vector to choose observations. 

```{r}

# (Let's give the academics their names back):

row.names(academics)[1] <- "Cosmo Kiveskoski"
row.names(academics)[2] <- "Kalja Naulapää"
row.names(academics)[3] <- "Miina Kenttä"
row.names(academics)[4] <- "Heppu Kaalinpää"
row.names(academics)[5] <- "Merikeiju Jumalankiitos"
row.names(academics)[6] <- "Meirami Pervonkallio"
```

If, for instance, we want to have a look at Miina's values, we can say:

```{r}
academics[ "Miina Kenttä" , ]

#If you don't like Miina, you can do:

academics[ row.names(academics) != "Miina Kenttä" , ]
```

Because the indexing implicitly creates logical vectors, you can also crate one explicitly and use it for choosing the observations:

```{r}
selection <- c(TRUE,TRUE,TRUE,TRUE,FALSE,FALSE)

academics[selection,]

# The inverse is defined as:

!selection

# so

academics[!selection,]

# For a quicker notation, ones and zeroes can also be used:

academics[as.logical(c(1,1,1,1,0,0)),]
```

As a final remark, indexing rows (observations) and columns (variables) can naturally be done simultaneously:

```{r}
academics[academics$gender == "F", 3:4]
```
